---
title: "MitoXplorer"
author: "Simon Haugaard"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required R libraries
```{r library}
if (!require("pacman")) install.packages("pacman")
pacman::p_load("edgeR")
pacman::p_load("readr")
pacman::p_load("readxl")
pacman::p_load("biomaRt")
pacman::p_load("magrittr")
pacman::p_load("tibble")
pacman::p_load("stringr")
pacman::p_load("ggplot2")
pacman::p_load("data.table")
pacman::p_load("ggplot2", "patchwork")
pacman::p_load("openxlsx")
pacman::p_load("dplyr")
pacman::p_load("data.table")
pacman::p_load("ggvenn")
pacman::p_load("pheatmap")
pacman::p_load_gh("altintasali/aamisc")

# install aamisc package for MDS and Volcano plots
#pacman::p_load("qvalue", "rain", "limma", "devtools")
#url <- "https://cran.r-project.org/src/contrib/Archive/HarmonicRegression/HarmonicRegression_1.0.tar.gz"
#pkgFile <- "HarmonicRegression_1.0.tar.gz"
#download.file(url = url, destfile = pkgFile)
#install.packages(pkgs=pkgFile, type="source", repos=NULL)
#file.remove(pkgFile)
#pacman::p_load_gh("altintasali/aamisc")

# Colours for publication
publication_colors <- c("AF" = "#285291", "Metformin" = "#7C1516", "Sham" = "#9B9B9B")

```

# Read data
## MitoXplorer3.0 web-data 
```{r read_data}
# Load count data from a local file and metadata through Excel
Mito_AF_vs_Sham_RA_RNA <- "../../../../RNA-seq/data/mitoXplorer/AF_vs_Sham_RA_RNA.txt"
Mito_AF_vs_Sham_RA_RNA <- fread(Mito_AF_vs_Sham_RA_RNA)

Mito_Metformin_vs_Placebo_RA_RNA <- "../../../../RNA-seq/data/mitoXplorer/Metformin_vs_placebo_RA_RNA.txt"
Mito_Metformin_vs_Placebo_RA_RNA <- fread(Mito_Metformin_vs_Placebo_RA_RNA)

# Classify genes for Metformin vs Placebo
Mito_Metformin_vs_Placebo_RA_RNA <- Mito_Metformin_vs_Placebo_RA_RNA %>%
  mutate(Significance = case_when(
    `P-VALUE` < 0.05 & LOG2FOLD > 0 ~ "Upregulated",
    `P-VALUE` < 0.05 & LOG2FOLD < 0 ~ "Downregulated",
    TRUE ~ "Non-significant"
  ))

# Classify genes for AF vs Sham
Mito_AF_vs_Sham_RA_RNA <- Mito_AF_vs_Sham_RA_RNA %>%
  mutate(Significance = case_when(
    `P-VALUE` < 0.05 & LOG2FOLD > 0 ~ "Upregulated",
    `P-VALUE` < 0.05 & LOG2FOLD < 0 ~ "Downregulated",
    TRUE ~ "Non-significant"
  ))
```

#Bar-Plot 
```{r fig.height=15, fig.width=20}
# Count genes for Metformin vs Placebo
count_Metformin_vs_Placebo <- Mito_Metformin_vs_Placebo_RA_RNA %>%
  group_by(PROCESS, Significance) %>%
  summarise(Count = n()) %>%
  ungroup()

# Count genes for AF vs Sham
count_AF_vs_Sham <- Mito_AF_vs_Sham_RA_RNA %>%
  group_by(PROCESS, Significance) %>%
  summarise(Count = n()) %>%
  ungroup()

# Combine the two datasets
combined_counts <- bind_rows(
  count_Metformin_vs_Placebo %>% mutate(Comparison = "Metformin vs Placebo"),
  count_AF_vs_Sham %>% mutate(Comparison = "AF vs Sham")
)

# Combine the counts of significant genes (upregulated + downregulated)
total_significant_Metformin_vs_Placebo <- count_Metformin_vs_Placebo %>%
  filter(Significance != "Non-significant") %>%
  group_by(PROCESS) %>%
  summarise(Total_Significant = sum(Count)) %>%
  arrange(desc(Total_Significant))

total_significant_AF_vs_Sham <- count_AF_vs_Sham %>%
  filter(Significance != "Non-significant") %>%
  group_by(PROCESS) %>%
  summarise(Total_Significant = sum(Count)) %>%
  arrange(desc(Total_Significant))

# For Metformin vs Placebo
count_Metformin_vs_Placebo <- count_Metformin_vs_Placebo %>%
  mutate(PROCESS = factor(PROCESS, levels = total_significant_Metformin_vs_Placebo$PROCESS))

# For AF vs Sham
count_AF_vs_Sham <- count_AF_vs_Sham %>%
  mutate(PROCESS = factor(PROCESS, levels = total_significant_AF_vs_Sham$PROCESS))

# Combine the two datasets
combined_counts <- bind_rows(
  count_Metformin_vs_Placebo %>% mutate(Comparison = "Metformin vs Placebo"),
  count_AF_vs_Sham %>% mutate(Comparison = "AF vs Sham")
)

# Plot with sorted facets
ggplot(combined_counts, aes(x = PROCESS, y = Count, fill = Significance)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ Comparison, scales = "free_x") +
  labs(title = "Gene Contributions to Biological Processes",
       x = "Biological Process", y = "Number of Genes", fill = "Regulation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top") +
  scale_fill_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Non-significant" = "grey"))



```
#Stacked Bar Plot
```{r fig.height=15, fig.width=20}
# Reorder the 'Significance' factor to control stacking order
combined_counts$Significance <- factor(combined_counts$Significance, levels = c("Upregulated", "Downregulated", "Non-significant"))

# Define colors to match your heatmap (red bottom, blue middle, grey top)
heatmap_colors <- c("Upregulated" = "#d73027", "Downregulated" = "#4575b4", "Non-significant" = "#d3d3d3")

ggplot(combined_counts, aes(x = PROCESS, y = Count, fill = Significance)) +
  geom_bar(stat = "identity", position = "stack") +  # Use 'stack' to stack the bars
  facet_wrap(~ Comparison, scales = "free_x", nrow = 1) +  # Arrange facets in a single row
  scale_fill_manual(values = heatmap_colors) +  # Apply heatmap-like colors
  labs(title = "Gene Contributions to Biological Processes",
       x = "Biological Process", y = "Number of Genes", fill = "Regulation") +
  theme_minimal(base_size = 16) +  # Increase base font size for readability
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),  # Increase x-axis text size and angle
    axis.text.y = element_text(size = 14),  # Increase y-axis text size
    axis.title.x = element_text(size = 18, face = "bold"),  # Larger and bold x-axis title
    axis.title.y = element_text(size = 18, face = "bold"),  # Larger and bold y-axis title
    plot.title = element_text(size = 24, face = "bold", hjust = 0.5),  # Larger and centered title
    legend.title = element_text(size = 18, face = "bold"),  # Larger legend title
    legend.text = element_text(size = 16),  # Larger legend text
    legend.position = "top",  # Position the legend at the top
    panel.grid.major = element_line(color = "grey80"),  # Soft grid lines for better visibility
    panel.grid.minor = element_blank(),  # Remove minor grid lines for a cleaner look
    strip.text = element_text(size = 18, face = "bold")  # Larger and bold facet labels
  )

```
#Dot plot
```{r fig.height=15, fig.width=20}
ggplot(combined_counts, aes(x = PROCESS, y = Significance, size = Count, color = Significance)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ Comparison, scales = "free_x", nrow = 1) +
  scale_size_continuous(range = c(5, 20)) +  # Increase the size range for points
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Non-significant" = "grey")) +
  labs(title = "Dot Plot of Significant Genes by Process",
       x = "Biological Process", y = "Regulation", size = "Number of Genes") +
  theme_minimal(base_size = 16) +  # Base font size increased for readability
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),  # Increase x-axis text size
    axis.text.y = element_text(size = 14),  # Increase y-axis text size
    axis.title.x = element_text(size = 18, face = "bold"),  # Larger and bold x-axis title
    axis.title.y = element_text(size = 18, face = "bold"),  # Larger and bold y-axis title
    plot.title = element_text(size = 24, face = "bold", hjust = 0.5),  # Larger and centered title
    legend.title = element_text(size = 18, face = "bold"),  # Larger legend title
    legend.text = element_text(size = 16),  # Larger legend text
    legend.position = "top",  # Keep legend at the top
    panel.grid.major = element_line(color = "grey80"),  # Soft grid lines for better visibility
    panel.grid.minor = element_blank(),  # Remove minor grid lines for a cleaner look
    strip.text = element_text(size = 18, face = "bold")  # Larger facet labels
  )

```

#Verdict 
  None of the generated graphs are as effective at conveying the message as the interactive plot available on the MitoXplorer web platform. 
  Therefore, I will use the MitoXplorer visualization for this purpose. However, to maintain consistency with the color scheme used throughout the publication, I will manually adjust the logFC values before uploading,   ensuring that red represents upregulation and blue represents downregulation.