---
title: "String Terminal"
author: "Simon Haugaard"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required R libraries
```{r library}
if (!require("pacman")) install.packages("pacman")
pacman::p_load("edgeR")
pacman::p_load("readr")
pacman::p_load("readxl")
pacman::p_load("biomaRt")
pacman::p_load("magrittr")
pacman::p_load("tibble")
pacman::p_load("stringr")
pacman::p_load("ggplot2")
pacman::p_load("data.table")
pacman::p_load("ggplot2", "patchwork")
pacman::p_load("openxlsx")
pacman::p_load("clusterProfiler")
library(dplyr)
library(missForest)
library(RColorBrewer)
library(ggplot2)
library(ggpubr)
library(tidyr)
library(SummarizedExperiment)
library(pheatmap)


```

#Read and Convert data
```{r}
# Define file paths
matrisome <- "../../../data/gene_annotation/Hs_Matrisome_Masterlist_downloaded_08_09_2024.csv"
dge_file <- "../../01_dge/output/dge_results.tsv.gz"
matrisome <- fread(matrisome)
dge <- fread(dge_file)
dge[, Direction := ifelse(logFC > 0, "up", "down")]


# Subset relevant contrasts and filter by P-value
relevant_contrasts <- c("AF_vs_Sham_RA", "AF_vs_Sham_LA", "Metformin_vs_AF_RA", "Metformin_vs_AF_LA", "AverageTreatmentEffect", "AverageDiseaseEffect")
dge <- dge[Contrast %in% relevant_contrasts]

# Ensure that the Gene Symbol column in matrisome data is clean and consistent
matrisome <- matrisome %>%
  dplyr::rename(GeneSymbol = `Gene Symbol`) %>%  # Rename for consistency
  dplyr::mutate(GeneSymbol = trimws(GeneSymbol))  # Remove any leading/trailing whitespace

# Merge dge with matrisome data to annotate ECM proteins
ecm_dge <- dge %>%
  dplyr::inner_join(matrisome, by = c("GeneName" = "GeneSymbol"))

# Save the annotated ECM data to a CSV file
write.csv(ecm_dge, file = "ECM_Annotated_DGE.csv", row.names = FALSE)
openxlsx::write.xlsx(x = ecm_dge, file = "../../../../Terminal/analysis/04_matrisome/output/ECM_Annotated_DGE.xlsx", asTable = TRUE)

# Subset to include only ECM core matrisome proteins
ecm_core <- ecm_dge %>% 
    dplyr::filter(`Matrisome Division` == "Core matrisome")

# Define significant matrisome genes
significant_matrisome_genes <- ecm_core$GeneName


```
#Overall look at matrisome between groups
```{r fig.width=20, fig.height=10, out.width='100%', echo=FALSE}
# Categorize proteins by significance
ecm_dge <- ecm_dge %>%
  mutate(Significance = case_when(
    adj.P.Val  < 0.05 & logFC > 0 ~ "Upregulated",
    adj.P.Val  < 0.05 & logFC < 0 ~ "Downregulated",
    TRUE ~ "Non-significant"
  ))
# Summarize the data by counting the number of proteins in each category and significance level
ecm_summary <- ecm_dge %>%
  group_by(`Matrisome Category`, Significance, Contrast) %>%
  summarise(Count = n()) %>%
  ungroup()

# Create the dot plot
ggplot(ecm_summary, aes(x = `Matrisome Category`, y = Significance, size = Count, color = Significance)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ Contrast, scales = "free_x", nrow = 1) +
  scale_size_continuous(range = c(3, 15)) +  # Adjust the size range for points based on count
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Non-significant" = "grey")) +
  labs(title = "Dot Plot of ECM Proteins by Category and Contrast",
       x = "Matrisome Category", y = "Regulation", size = "Number of Proteins") +
  theme_minimal(base_size = 16) +  # Increase base font size for readability
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),  # Increase x-axis text size and angle
    axis.text.y = element_text(size = 14),  # Increase y-axis text size
    axis.title.x = element_text(size = 18, face = "bold"),  # Larger and bold x-axis title
    axis.title.y = element_text(size = 18, face = "bold"),  # Larger and bold y-axis title
    plot.title = element_text(size = 24, face = "bold", hjust = 0.5),  # Larger and centered title
    legend.title = element_text(size = 18, face = "bold"),  # Larger legend title
    legend.text = element_text(size = 16),  # Larger legend text
    legend.position = "top",  # Position the legend at the top
    panel.grid.major = element_line(color = "grey80"),  # Soft grid lines for better visibility
    panel.grid.minor = element_blank(),  # Remove minor grid lines for a cleaner look
    strip.text = element_text(size = 18, face = "bold")  # Larger and bold facet labels
  )

# Ensure the summarise warning doesn't affect your plot
ecm_summary <- ecm_dge %>%
  mutate(Significance = case_when(
    adj.P.Val  < 0.05 & logFC > 0 ~ "Upregulated",
    adj.P.Val  < 0.05 & logFC < 0 ~ "Downregulated",
    TRUE ~ "Non-significant"
  )) %>%
  group_by(`Matrisome Category`, Significance, Contrast) %>%
  summarise(Count = n(), .groups = 'drop')  # Include '.groups = 'drop'' to avoid the warning

# Create the dot plot
ggplot(ecm_summary, aes(x = `Matrisome Category`, y = Significance, size = Count, color = Significance)) +
  geom_point(alpha = 0.8) +  # Slightly increase the opacity for better visibility
  facet_wrap(~ Contrast, scales = "free_x", nrow = 1) +
  scale_size_continuous(range = c(3, 20), guide = guide_legend(title = "Number of Proteins")) +  # Larger size range for points
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Non-significant" = "grey")) +  # Color palette
  labs(title = "Differential Regulation of ECM Proteins by Category and Contrast",
       x = "Matrisome Category", y = "Regulation Status") +
  theme_pubr(base_size = 18, border = TRUE) +  # Use ggpubr's theme for a clean look
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 16, face = "bold"),  # Bold and larger x-axis text
    axis.text.y = element_text(size = 16, face = "bold"),  # Bold and larger y-axis text
    axis.title.x = element_text(size = 20, face = "bold"),  # Larger and bold x-axis title
    axis.title.y = element_text(size = 20, face = "bold"),  # Larger and bold y-axis title
    plot.title = element_text(size = 26, face = "bold", hjust = 0.5),  # Larger and centered title
    legend.title = element_text(size = 18, face = "bold"),  # Larger legend title
    legend.text = element_text(size = 16),  # Larger legend text
    legend.position = "top",  # Keep legend at the top
    panel.grid.major = element_line(color = "grey90"),  # Softer grid lines for a cleaner look
    panel.grid.minor = element_blank(),  # Remove minor grid lines
    strip.text = element_text(size = 18, face = "bold")  # Larger and bold facet labels
  )

# Extract significant upregulated and downregulated proteins for each contrast
ecm_dge <- ecm_dge %>%
  mutate(Significance = case_when(
    adj.P.Val  < 0.05 & logFC > 0 ~ "Upregulated",
    adj.P.Val  < 0.05 & logFC < 0 ~ "Downregulated",
    TRUE ~ "Non-significant"
  ))

# Extract significant upregulated proteins with Matrisome Category
significant_up <- ecm_dge %>%
  filter(Significance == "Upregulated") %>%
  group_by(Contrast, `Matrisome Category`) %>%
  summarise(Upregulated_Proteins = paste(GeneName, collapse = ", "), .groups = 'drop')

# Extract significant downregulated proteins with Matrisome Category
significant_down <- ecm_dge %>%
  filter(Significance == "Downregulated") %>%
  group_by(Contrast, `Matrisome Category`) %>%
  summarise(Downregulated_Proteins = paste(GeneName, collapse = ", "), .groups = 'drop')

# Print out the results
cat("Significant Upregulated Proteins:\n")
print(significant_up)

cat("\nSignificant Downregulated Proteins:\n")
print(significant_down)

```


#Seperate plots for publication
```{r fig.width=10, fig.height=10, out.width='100%', echo=FALSE}
# Filter for LA contrasts
ecm_summary_LA <- ecm_summary %>%
  filter(grepl("_LA", Contrast))

# Filter for RA contrasts
ecm_summary_RA <- ecm_summary %>%
  filter(grepl("_RA", Contrast))

# Filter for Average contrasts
ecm_summary_Average <- ecm_summary %>%
  filter(grepl("Average", Contrast))

ggplot(ecm_summary_LA, aes(x = `Matrisome Category`, y = Significance, size = Count, color = Significance)) +
  geom_point(alpha = 0.8) +
  facet_wrap(~ Contrast, scales = "free_x", nrow = 1) +
  scale_size_continuous(range = c(3, 20), guide = guide_legend(title = "Number of Proteins")) +
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Non-significant" = "grey")) +
  labs(title = "Differential Regulation of ECM Proteins in LA by Category and Contrast",
       x = "Matrisome Category", y = "Regulation Status") +
  theme_pubr(base_size = 18, border = TRUE) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 16, face = "bold"),
    axis.text.y = element_text(size = 16, face = "bold"),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    plot.title = element_text(size = 26, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 18, face = "bold"),
    legend.text = element_text(size = 16),
    legend.position = "top",
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 18, face = "bold")
  )

ggplot(ecm_summary_RA, aes(x = `Matrisome Category`, y = Significance, size = Count, color = Significance)) +
  geom_point(alpha = 0.8) +
  facet_wrap(~ Contrast, scales = "free_x", nrow = 1) +
  scale_size_continuous(range = c(3, 20), guide = guide_legend(title = "Number of Proteins")) +
  scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue", "Non-significant" = "grey")) +
  labs(title = "Differential Regulation of ECM Proteins in RA by Category and Contrast",
       x = "Matrisome Category", y = "Regulation Status") +
  theme_pubr(base_size = 18, border = TRUE) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 16, face = "bold"),
    axis.text.y = element_text(size = 16, face = "bold"),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    plot.title = element_text(size = 26, face = "bold", hjust = 0.5),
    legend.title = element_text(size = 18, face = "bold"),
    legend.text = element_text(size = 16),
    legend.position = "top",
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 18, face = "bold")
  )

```

# Enrichment analysis 
```{r read enrichment files}
# Add a new column to categorize upregulated and downregulated genes based on log fold-change (logFC)
dge[, Direction := ifelse(logFC > 0, "up", "down")]

# Read gene annotation data from a specified file path
geneinfo_file <- "../../../data/gene_annotation/horse_gene_annotation.tsv.gz"
geneinfo <- fread(geneinfo_file)

# Set new column names for better clarity
setnames(geneinfo, new = c("ENSEMBL", "ENSEMBLv", "Description_detailed", 
                           "Chr", "Start", "End", "Strand", "GENENAME", "ENTREZID", "Description"))

# Remove unwanted columns and duplicated entries based on ENSEMBL IDs
geneinfo <- geneinfo[, c("ENSEMBLv", "Description_detailed") := NULL]  # Remove columns not needed
geneinfo <- geneinfo[!duplicated(ENSEMBL), ]                           # Retain only unique entries

# Extract gene names (GENENAME) from DGE data, assuming "GeneName" column exists in DGE file
dge[, GENENAME := sub(".*_", "", GeneName)]

# Merge DGE data with gene annotation data based on GENENAME
# The merge is performed to align DGE results with additional annotation information
dge <- merge(dge, geneinfo[, .(ENSEMBL, GENENAME, ENTREZID)], by = "GENENAME", all.x = TRUE)
dge <- dge[!is.na(ENSEMBL), ]

# Reorder columns to prioritize key identifiers (ENSEMBL, GENENAME, ENTREZID)
setcolorder(dge, c("ENSEMBL", "GENENAME", "ENTREZID", setdiff(names(dge), c("ENSEMBL", "GENENAME", "ENTREZID"))))

# Convert ENTREZID column to character for downstream compatibility
dge[, ENTREZID := as.character(ENTREZID)]

# Filter the DGE results for specific relevant contrasts of interest
dge <- dge[Contrast %in% c("AF_vs_Sham_RA", "AF_vs_Sham_LA", 
                           "Metformin_vs_AF_RA", "Metformin_vs_AF_LA", 
                           "AverageTreatmentEffect", "AverageDiseaseEffect")]

# Keep only entries with valid Entrez IDs for enrichment analysis
dge_entrez <- dge[!is.na(ENTREZID),]

# Define statistical cutoffs for differential gene expression and enrichment analysis
dge_cut <- 0.05        # Significance cutoff for differential gene expression (p-value)
enrich_cut <- 0.05     # Significance cutoff for enrichment analysis (q-value)


```

The analysis will be performed according to following statistical cutoffs.

- *Differential gene expression:* p-value < `r dge_cut`
- *Enrichment analysis:* adjusted p-value (Q-value) < `r enrich_cut`

# Matrisome Enrichment
```{r matrisome}
# Read & Clean Matrisome data
matrisome <- "../../../data/gene_annotation/Hs_Matrisome_Masterlist_downloaded_08_09_2024.csv"
matrisome <- fread(matrisome)
matrisome <- matrisome %>%
  dplyr::rename(GeneSymbol = `Gene Symbol`) %>%  # Rename for consistency
  dplyr::mutate(GeneSymbol = trimws(GeneSymbol))  # Remove any leading/trailing whitespace

# Map GeneSymbol to ENSEMBL using geneinfo
matrisome_ensembl <- merge(matrisome, geneinfo, by.x = "GeneSymbol", by.y = "GENENAME", all.x = TRUE)
matrisome_ensembl <- matrisome_ensembl[!is.na(ENSEMBL)]

# Create TERM2GENE
term2gene_matrisome <- matrisome_ensembl[, .(`Matrisome Category`, ENSEMBL)]

# Initialize matrisome_gsea list
matrisome_gsea <- list()

# Split differential gene expression (DGE) data into separate lists by contrast
dge_split <- split(dge, f = dge$Contrast)

# Loop through DGE contrasts
for (contrast in names(dge_split)) {
    message(paste0("Running GSEA for contrast '", contrast, "'"))

    # Prepare geneList
    geneList <- dge_split[[contrast]]$logFC
    names(geneList) <- dge_split[[contrast]]$ENSEMBL
    geneList <- sort(geneList, decreasing = TRUE)

    # Run GSEA
    gsea_result <- GSEA(
        geneList = geneList, 
        TERM2GENE = term2gene_matrisome, 
        TERM2NAME = NULL,  # Not required since categories are already descriptive
        pvalueCutoff = 0.05,
        minGSSize = 10, 
        maxGSSize = 500, 
        eps = 0, 
        nPermSimple = 10000
    )@result

    # Add metadata and save results
    if (nrow(gsea_result) > 0) {
        gsea_result$Contrast <- contrast
    } else {
        gsea_result <- data.frame(Contrast = character(), Database = character())
    }
    matrisome_gsea[[contrast]] <- gsea_result
}

# Combine all results
matrisome_gsea_res <- do.call(rbind, matrisome_gsea) %>% setDT
matrisome_gsea_res[, direction := ifelse(NES < 0, "down", "up")]

# Split results by contrast
matrisome_gsea_res_split <- split(matrisome_gsea_res, f = matrisome_gsea_res$Contrast)

# Save results
openxlsx::write.xlsx(matrisome_gsea_res_split, file = "../output/matrisome_gsea_results.xlsx", asTable = TRUE)
fwrite(matrisome_gsea_res, file = "../output/matrisome_gsea_results.tsv.gz", sep = "\t")

# Define contrasts for plotting
# Filter the contrasts for the combined plot
selected_contrasts <- c("AF_vs_Sham_RA", "AF_vs_Sham_LA", "Metformin_vs_AF_RA", "Metformin_vs_AF_LA")

# Subset results for selected contrasts
combined_results <- matrisome_gsea_res[Contrast %in% selected_contrasts]

# Plotting function for combined visualization
# Function to plot all contrasts in a single dot plot
plot_combined_dotplot <- function(data_split, contrasts, topn = 10, topn.pref = "qval", qcut = enrich_cut, nchar = 60,
                                  direction = "direction", group = "Contrast", dot = "NES", qval = "qvalue", 
                                  term.id = "ID", term.name = "Description") {
  
  # Combine all selected contrasts into one data.table
  combined_data <- rbindlist(lapply(contrasts, function(contrast) {
    if (contrast %in% names(data_split)) {
      filtered_data <- data_split[[contrast]]
      if (nrow(filtered_data) > 0) {
        # Add the contrast name to the data
        filtered_data[, Contrast := contrast]
        return(filtered_data)
      }
    }
    NULL
  }), use.names = TRUE, fill = TRUE)
  
  # Check if data is available
  if (is.null(combined_data) || nrow(combined_data) == 0) {
    message("No data available for the selected contrasts.")
    return(NULL)
  }
  
  # Filter for top terms per contrast
  top_terms <- combined_data[, .SD[order(get(topn.pref))[1:topn]], by = Contrast]
  
  # Plot using aamisc::dotplotEnrich
  return(aamisc::dotplotEnrich(dt = top_terms, 
                               topn = topn, 
                               topn.pref = topn.pref, 
                               qcut = qcut, 
                               nchar = nchar, 
                               direction = direction, 
                               group = group, 
                               dot = dot, 
                               qval = qval, 
                               term.id = term.id, 
                               term.name = term.name))
  
  
}

# Plot the combined results
plot_combined_dotplot(data_split = matrisome_gsea_res_split, contrasts = selected_contrasts)


# Combine all selected contrasts into one data.table for inspection
combined_data <- rbindlist(lapply(selected_contrasts, function(contrast) {
  if (contrast %in% names(matrisome_gsea_res_split)) {
    filtered_data <- matrisome_gsea_res_split[[contrast]]
    if (nrow(filtered_data) > 0) {
      # Add the contrast name to the data
      filtered_data[, Contrast := contrast]
      return(filtered_data)
    }
  }
  NULL
}), use.names = TRUE, fill = TRUE)


# Select the top 5 terms per contrast and direction
topn <- 5
filtered_data <- combined_data[, .SD[order(qvalue)][1:topn], by = .(Contrast, direction)]
filtered_data <- filtered_data[!is.na(ID) & !is.na(Description)]

# Check the filtered data
print(filtered_data)

plot_combined_dotplot <- function(filtered_data, topn = 5, topn.pref = "qvalue", qcut = 0.05, nchar = 60,
                                  direction = "direction", group = "Contrast", dot = "NES", qval = "qvalue", 
                                  term.id = "ID", term.name = "Description") {
  
  # Check if topn.pref is valid
  if (!topn.pref %in% names(filtered_data)) {
    stop(paste("Column", topn.pref, "not found in data. Available columns:", paste(names(filtered_data), collapse = ", ")))
  }
  
  # Plot using aamisc::dotplotEnrich
  return(aamisc::dotplotEnrich(dt = filtered_data, 
                               topn = topn,  # Use the topn parameter
                               topn.pref = topn.pref,  # Explicitly set topn.pref
                               qcut = qcut, 
                               nchar = nchar, 
                               direction = direction, 
                               group = group, 
                               dot = dot, 
                               qval = qval, 
                               term.id = term.id, 
                               term.name = term.name))
}

# Clean filtered data
filtered_data <- filtered_data[!is.na(ID) & !is.na(Description)]
filtered_data[, qval := qvalue]

# Generate the plot
combined_plot <- plot_combined_dotplot(filtered_data, topn = 5, topn.pref = "qval")
print(combined_plot)
ggsave("../output/combined_dotplot_hq.png", plot = combined_plot, width = 12, height = 8, dpi = 600)


```


# Session Info
```{r session_info}
sessionInfo()
```

